Using Cell Phone Keyboards is (\NP) Hard

### 2010-01-22 article.tex
Proven?

The baseline we compare against is the most painful method of text entry.  I
imagine that no ``digital native'' actually uses this input method, and that it
is available on all phones solely that their parents might use it.  In this
method, to type an `a', the user of the cellphone types the `2' key; to type a
`b' they type `22', to type `c' they type `222', to type a `d' they type `3',
to type an `e' they type `33', and so on.  To type a word with multiple
letters that use the same key, such as `accept', the user must pause
between keystrokes.  Thus, the full key entry sequence for `accept', using
`.' to indicate a pause, is `2.22.223378'.

@@

In an effort to minimize the pain of cellphone keyboard typing, cellular
telephone manufacturers have created the T9 input method, which attempts to
guess which letter (of the 3 or 4 possible) is intended when a user hits a
single key.  Enhancements of this input method also provide
\newword{speculative lookahead} to report, at any given time, what word it is
that the user is most likely trying to enter and provide a completion.
Unfortunately, because there are fewer keys on the cellphone keyboard than
there are letters in the English alphabet, there are words which have different
spellings but the same input sequence.  As an example, in the traditional
mapping of keys to characters both ``me'' and ``of'' have the input sequence
``63''.  Two words with the same input sequence force the user to press a third
button to cycle through the possibilities in order from most likely (``of'') to
least likely (``me'').  Even worse: ``home'', ``good'', ``gone'', ``hood'',
``hoof'', ``hone'', and ``goof'' all have the input sequence ``4663''.  If
the * key is used as the ``next match'' button, then the user will actually
have to type in ``4663******'' to type in the word ``goof'', for a total of 10
button presses --- exactly the same number of button presses required using the
default layout and the basic input method, and three more button presses than
is required when using the basic input method with an optimized layout.

@@

\Question\ What is the partion $P$ of $A$ into $k$ sets which minimizes the total button presses required to enter the entire corpus using the T9 input method?
\end{prob}

@@

\begin{prob}~\\
\Instance\ An alphabet $A$, a set of words and their associated frequencies $W$, a number of keys $k$ ($|A| > k$), and a number $t$.

\Question\ Is there a partition of $A$ into $k$ sets in which the total number of keystrokes required to enter the entire corpus using the T9 input method requires no more than $t$ button presses?
\end{prob}

@@

\begin{proof}[{\sc UniquePathColoring} is \NP-complete]
To prove that {\sc UniquePathColoring} is \NP-complete, we begin by noting that any coloring of $V$ may be verified, in polynomial time, to map each path to a unique sequence of colors.  Therefore, the problem is in \NP.  To prove completeness, we reduce from {\sc GraphColoring}\cite{gandj}.

@@

Now we prove that given a $k$-unique-path coloring of $G'$ and $P$ we can create a $k$-coloring of $G$.
\qed
\end{proof}

>>>>

The baseline we compare against is the most painful method of text entry.  In this
method, to type an `a', the user of the cellphone types the `2' key; to type a
`b' they type `22', to type `c' they type `222', to type a `d' they type `3',
to type an `e' they type `33', and so on.  To type a word with multiple letters
that use the same key, such as `accept', the user must pause between
keystrokes.  Thus, the full key entry sequence for `accept', using `.' to
indicate a pause, is `2.22.223378'.

@@

In an effort to minimize the pain of cellphone keyboard typing, cellular
telephone manufacturers have created the T9 input method, which attempts to
guess which letter (of the 3 or 4 possible) is intended when a user hits a
single key.  Enhancements of this input method also provide
\newword{speculative lookahead} to report, at any given time, what word it is
that the user is most likely trying to enter and provide a completion.
Unfortunately, because there are fewer keys on the cellphone keyboard than
there are letters in the English alphabet, there are words which have different
spellings but the same input sequence.  As an example, in the traditional
mapping of keys to characters both `me' and `of' have the input sequence
`63'.  Two words with the same input sequence force the user to press a third
button to cycle through the possibilities in order from most likely (`of') to
least likely (`me').  Even worse: `home', `good', `gone', `hood',
`hoof', `hone', and `goof' all have the input sequence `4663'.  If
the * key is used as the ``next match'' button, then the user will actually
have to type in `4663******' to type in the word `goof', for a total of 10
button presses --- exactly the same number of button presses required using the
default layout and the basic input method, and three more button presses than
is required when using the basic input method with an optimized layout.

@@

\Question\ What is the partition $P$ of $A$ into $k$ sets which minimizes the
total button presses required to enter the entire corpus using the T9 input
method?
\end{prob}

@@

\begin{prob}~\\
\label{minstrokesdecision}
\Instance\ A set $A$, a set $W$ of sequences of elements of $A$, a mapping $f$ from sequences in $W$ to the integers, and a number $t$.

\Question\ Is there a partition $P$ of $A$ into $k$ sets such that, if we denote $P(w)$ as the sequence of partitions $[p_i | w_i \in p_i, p_i \in P]$,
$$\sum_{w\in W} (\mathrm{len}(w)+\mathrm{order}(w,P,W,f)) * f(w) \le t$$
where len$(w)$ is the length of the sequence $w$ and order$(w,P,W,f)$ is the size of $\left\{s\in W | P(s) = P(w) \land f(s) > f(w)\right\}$, which is the set of sequences which map to the same sequence of partitions, but are strictly more popular.
\end{prob}

@@

\begin{theorem}{\sc UniquePathColoring} is \NP-complete.\end{theorem}
\begin{proof}
To prove that {\sc UniquePathColoring} is \NP-complete, we begin by noting that
any coloring of $V$ may be verified, in polynomial time, to map each path to a
unique sequence of colors.  Therefore, the problem is in \NP.  To prove
completeness, we reduce from {\sc GraphColoring}\cite{gandj}.

@@

Now we prove that given a $k$-unique-path coloring of $G'$ and $P$ we can
create a $k$-coloring of $G$.  Given a $k$-unique path coloring, we are
guaranteed that no path in $P$ has the exact same coloring as any other path in
$P$.  This means that, for all $i$, the two paths in $p_i$ are distinct.  The
only difference between the two paths in $p_i$ is in the first vertex of the
path.  Therefore, for every $p_i$, the vertices of corresponding edge in must
be given distinct colors, and this is true for all $i$ and $e \in E$.
Therefore, for all $(u,v)\in E$, the color assigned to $u$ is distinct from the
color assigned to $v$.  Therefore, from a $k$-unique-path coloring of $G'$ and
$P$, we have a $k$-coloring of $G=(V,E)$.

\qed 
\end{proof}

Now that we have proven {\sc UniquePathColoring} to be \NP-complete, we
immediately use it in a reduction.  
\begin{theorem}{\sc MinimumT9Keystrokes} (the decision problem form,
Problem~\ref{minstrokesdecision}) is \NP-complete.  \end{theorem}
